---
title: "Tongue root advancement and vowel duration: a gradient effect?"
author: "Stefano Coretta"
date: "The University of Manchester"
institute: "13 April, BAAP 2018 (Canterbury)"
output:
  beamer_presentation:
    citation_package: natbib
    highlight: zenburn
    latex_engine: xelatex
fontsize: 12pt
bibliography: linguistics.bib
biblio-style: unified.bst
header-includes:
- \frenchspacing
- \usepackage{cleveref}
- \usetheme{metropolis}
- \titlegraphic{\includegraphics[height=1cm]{./graphics/logo.png}}
- \usecolortheme[snowy]{owl}
nocite: |
  @westbury1983, @davidson2016, @abramson2017, @westbury1983, @ohala2011, @ahn2015
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
theme_set(theme_minimal(base_size = 20))
library(tidymv) # devtools::install_github("stefanocoretta/tidymv@e277a70", build_vignettes = TRUE)
library(rticulate) # devtools::install_github("stefanocoretta/rticulate@bcef263", build_vignettes = TRUE)
library(lme4)
library(effects)
library(itsadug)
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r read-data, include=FALSE}
speakers <- read_csv("./datasets/speakers.csv")
stimuli <- read_csv("./datasets/stimuli.csv")

vowels <- list.files(
  path = "./datasets/acoustics/",
  pattern = "*-durations.csv",
  full.names = TRUE
) %>%
  map_df(~read_csv(., na = "--undefined--")) %>%
  left_join(y = speakers) %>%
  left_join(y = stimuli) %>%
  mutate(c2_phonation = factor(c2_phonation, levels = c("voiceless", "voiced"))) %>%
  mutate_if(is.character, as.factor)

duration <- list.files(
  path = "./datasets/acoustics/",
  pattern = "*-durations.csv",
  full.names = TRUE
) %>%
  map_df(~read_csv(., na = "--undefined--")) %>%
  select(-speaker)

columns <- c(
  "speaker",
  "seconds",
  "rec_date",
  "prompt",
  "label",
  "TT_displacement_sm",
  "TT_velocity",
  "TT_velocity_abs",
  "TD_displacement_sm",
  "TD_velocity",
  "TD_velocity_abs",
  "TR_displacement_sm",
  "TR_velocity",
  "TR_velocity_abs"
)

tra_series <- list.files(
  path = "./datasets/ultrasound/",
  pattern = "*-vowel-series.tsv",
  full.names = TRUE
) %>%
  read_aaa(., columns, format = "wide") %>%
  select(-(X_1:Y_42)) %>%
  left_join(duration) %>%
  mutate(word = word(prompt, 2)) %>%
  left_join(y = speakers) %>%
  left_join(y = stimuli) %>%
  mutate(
    c2_phonation = ordered(c2_phonation, levels = c("voiceless", "voiced")),
    vowel = ordered(vowel, levels = c("a", "o", "u")),
    c2_place = ordered(c2_place, levels = c("coronal", "velar")),
    proportion = (seconds - v_onset) / (v_offset - v_onset),
    tra = -TR_displacement_sm
  ) %>%
  group_by(speaker) %>%
  mutate(tra_z = scale(tra)) %>%
  ungroup() %>%
  mutate_if(is.character, as.factor) %>%
  mutate(
    speaker_contexts = interaction(speaker, c2_place, vowel),
    place_vowel = interaction(c2_place, vowel),
    place_vowel_phon = interaction(c2_place, vowel, c2_phonation)
  )

contrasts(tra_series$c2_phonation) <- "contr.treatment"
contrasts(tra_series$vowel) <- "contr.treatment"
contrasts(tra_series$c2_place) <- "contr.treatment"
contrasts(tra_series$place_vowel) <- "contr.treatment"
contrasts(tra_series$place_vowel_phon) <- "contr.treatment"

columns_2 <- c(
  "speaker",
  "seconds",
  "rec_date",
  "prompt",
  "label",
  "TT_displacement_sm",
  "TT_velocity",
  "TT_velocity_abs",
  "TD_displacement_sm",
  "TD_velocity",
  "TD_velocity_abs"
)

splines <- list.files(
  path = "./datasets/ultrasound",
  pattern = "*-tongue-cart.tsv",
  full.names = TRUE
) %>%
  map_df(~read_aaa(., columns_2)) %>%
  mutate(word = word(prompt, 2)) %>%
  left_join(y = speakers) %>%
  left_join(y = stimuli) %>%
  mutate(
    c2_phonation = ordered(c2_phonation, levels = c("voiceless", "voiced")),
    vowel = ordered(vowel, levels = c("a", "o", "u")),
    c2_place = ordered(c2_place, levels = c("coronal", "velar")),
    vow_phon = as.ordered(interaction(vowel, c2_phonation)),
    c2_place_phon = as.ordered(interaction(c2_place, c2_phonation))
  ) %>%
  mutate_if(is.character, as.factor) %>%
  arrange(rec_date, fan_line) %>%
  create_event_start("rec_date")

contrasts(splines$c2_phonation) <- "contr.treatment"
contrasts(splines$vowel) <- "contr.treatment"
contrasts(splines$c2_place) <- "contr.treatment"
contrasts(splines$vow_phon) <- "contr.treatment"
contrasts(splines$c2_place_phon) <- "contr.treatment"

```

# Background

<!--
* **Correlates of voicing**
    * *shorter VOT* [@westbury1983;@davidson2016;@abramson2017]
    * *tongue root advancement TRA* [@westbury1983;@ohala2011]
    * *correlation VOT ~ TRA* [@ahn2015]
    * *longer vowel duration* [@house1953;@peterson1960;@chen1970;@klatt1973;@lisker1974;@fowler1992;@lampp2004]

* **Relation between vowel duration and TRA**
-->

\begin{figure}
  \includegraphics[width=0.8\textwidth]{./graphics/voicing_0.pdf}
\end{figure}

# Background

\begin{figure}
  \includegraphics[width=0.8\textwidth]{./graphics/voicing_1.pdf}
\end{figure}

# Background

\begin{figure}
  \includegraphics[width=0.8\textwidth]{./graphics/voicing_2.pdf}
\end{figure}

# Background

\begin{figure}
  \includegraphics[width=0.8\textwidth]{./graphics/voicing_3.pdf}
\end{figure}

# Background

\begin{figure}
  \includegraphics[width=0.8\textwidth]{./graphics/voicing_4.pdf}
\end{figure}

# Background

\begin{figure}
  \includegraphics[width=0.8\textwidth]{./graphics/voicing_5.pdf}
\end{figure}

# Background

* **Voicing effect** (VE): **vowels are longer when followed by voiced stops** [@house1953;@peterson1960;@chen1970;@klatt1973;@lisker1974;@fowler1992;@lampp2004]
    * **Italian**: voicing effect of 35 msec [@farnetani1986]
    * **Polish**: mixed results
        * @keating1984: no effect
        * @nowak2006 PhD dissertation: 4.5 msec effect
* **Larger study**: relative timing of laryngeal and lingual activity
    * Simultaneous UTI + EGG + audio
* **This study**: exploratory, data driven

# Methods (a summary)

* **Participants**: 4 Italians (2 F, 2 M), 4 Polish (2 F, 2 M)
* **Targets**
    * \textbf{C\textsubscript{1}V\textsubscript{1}C\textsubscript{2}V\textsubscript{1}}
    * C\textsubscript{1} = /p/, V\textsubscript{1} = /a, o, u/, C\textsubscript{2} = /t, d, k, g/
    * *pata*, *pada*, *paka*, ..., *poto*, *podo*, ...
* **Frame sentence**
    - *Dico X lentamente*, 'I say X slowly'
    - *Mówię X teraz*, 'I say X now'
* **Data**
    * Durational data from acoustics
    * Tongue contours from ultrasound tongue imaging
* **Reproducibility**
    - <https://github.com/stefanocoretta/2018-baap>

# Results: Vowel duration

```{r vowel-duration-plot, echo=FALSE}
vowels %>%
  group_by(speaker) %>%
  mutate(vowel_duration = scale(vowel_duration)) %>%
  ggplot(aes(vowel, vowel_duration, fill = c2_phonation)) +
  geom_boxplot() +
  facet_grid(~ language, labeller = labeller(language = c(italian = "Italian", polish = "Polish"))) +
  xlab("Vowel") +
  ylab("Vowel duration (scaled)") +
  scale_fill_manual(name = "C2 voicing", values = cbPalette[2:3])
```

# Results: Vowel duration

```{r vow-it-lm, include=FALSE}
vowels_it <- filter(vowels, language == "italian")

vowels_it_lm <- lmer(
  vowel_duration ~
    c2_phonation +
    c2_place +
    vowel +
    c2_phonation:vowel +
    sentence_duration +
    (1+c2_phonation|speaker) +
    (1|word),
  data = vowels_it,
  REML = FALSE
)

vowels_it_lm_null <- lmer(
  vowel_duration ~
#    c2_phonation +
    c2_place +
    vowel +
#    c2_phonation:vowel +
    sentence_duration +
    (1+c2_phonation|speaker) +
    (1|word),
  data = vowels_it,
  REML = FALSE
)

summary(vowels_it_lm)$coeff

vowels_it_mc <- anova(vowels_it_lm_null, vowels_it_lm)
```

```{r vow-pl-lm, include=FALSE}
vowels_pl <- filter(vowels, language == "polish")

vowels_pl_lm <- lmer(
  vowel_duration ~
    c2_phonation +
    c2_place +
    vowel +
    c2_phonation:vowel +
    c2_place:vowel +
    sentence_duration +
    (1+c2_phonation|speaker) +
    (1|word),
  data = vowels_pl,
  REML = FALSE
)

vowels_pl_lm_null <- lmer(
  vowel_duration ~
#    c2_phonation +
    c2_place +
    vowel +
#    c2_phonation:vowel +
    c2_place:vowel +
    sentence_duration +
    (1+c2_phonation|speaker) +
    (1|word),
  data = vowels_pl,
  REML = FALSE
)

summary(vowels_pl_lm)$coeff

vowels_pl_mc <- anova(vowels_pl_lm_null, vowels_pl_lm)
```

```{r vow-lm-effects, include=FALSE}
vow_it_effects <- as_tibble(
  effect("c2_phonation", vowels_it_lm)
) %>%
  mutate(language = "Italian")

vow_pl_effects <- as_tibble(
  effect("c2_phonation", vowels_pl_lm)
) %>%
  mutate(language = "Polish")

vow_effects <- bind_rows(vow_it_effects, vow_pl_effects)
```

* Linear mixed-effects models [@bates2015;@kuznetsova2016]

* **Italian**: $\beta$ = `r round(summary(vowels_it_lm)$coeff[2, 1])` msec, $\chi^2$(`r vowels_it_mc[2,7]`) = `r round(vowels_it_mc[2,6], 2)`, p = `r vowels_it_mc[2,8]`

* **Polish**: $\beta$ = `r round(summary(vowels_pl_lm)$coeff[2, 1])` msec, $\chi^2$(`r vowels_pl_mc[2,7]`) = `r round(vowels_pl_mc[2,6], 2)`, p = `r vowels_pl_mc[2,8]`

```{r vow-lm-effects-plot, include=FALSE}
ggplot(vow_effects, aes(c2_phonation, fit)) +
  geom_errorbar(aes(ymin = fit - se, ymax = fit + se), width = 0.2, colour = cbPalette[2]) +
  geom_point() +
  facet_grid(~ language, scales = "free")
```

# Results: Vowel duration

```{r vowels-lm-coef}
as_tibble(rbind(coef(vowels_it_lm)$speaker[2], coef(vowels_pl_lm)$speaker[2]), rownames = "speaker") %>%
  mutate(language = c(rep("Italian", 4), rep("Polish", 4))) %>%
  ggplot(aes(speaker, c2_phonationvoiced, fill = language)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(name = "Language", values = cbPalette[2:3]) +
  labs(title = "By-speaker random coefficients", y = "msec")
```

# Results: Tongue contours

* **Midsagittal tongue contours**
    * From *within consonant closure* [at maximum tongue displacement, @strycharczuk2015], polar coordinates [@heyne2015a;@heyne2015;@mielke2015]

* **Generalised additive mixed models (GAMMs)** [@wood2006;@soskuthy2017;@van-rij2017]

* **Polar GAMMs** with the `rticulate` R package [@coretta2018;@coretta2018a]

* **General trends**
    * Idiosyncratic use of TRA
    * 2 speakers with relatively greater TRA

# Results: Tongue contours

\begin{figure}
  \includegraphics[width = \textwidth]{./graphics/it01-uti.pdf}
\end{figure}

```{r it01, eval=FALSE, include=FALSE}
it01 <- filter(splines, speaker == "it01", label %in% c("max_TT", "max_TD"))

it01_polar_gam <- polar_gam(
  Y ~
    c2_phonation +
    vowel +
    c2_place +
#    vow_phon +
#    c2_place_phon +
    s(X, bs = "cr") +
    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
#    s(X, by = vow_phon, bs = "cr") +
#    s(X, by = c2_place_phon, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = it01,
  method = "ML"
)

# No need for interactions

rho <- start_value_rho(it01_polar_gam)

it01_polar_gam_ar <- polar_gam(
  Y ~
    c2_phonation +
    vowel +
    c2_place +
    s(X, bs = "cr") +
    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = it01,
  method = "ML",
  rho = rho,
  AR_start = it01$start.event
)

it01_polar_gam_ar_null <- polar_gam(
  Y ~
#    c2_phonation +
    vowel +
    c2_place +
    s(X, bs = "cr") +
#    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = it01,
  method = "ML",
  rho = rho,
  AR_start = it01$start.event
)

compareML(it01_polar_gam_ar_null, it01_polar_gam_ar)

acf_resid(it01_polar_gam_ar, split_pred = "AR.start")

plot_polar_smooths(
  it01_polar_gam_ar,
  X,
  c2_phonation,
  facet_terms = c2_place + vowel
) +
  labs(title = "IT01")
```

# Results: Tongue contours

\begin{figure}
  \includegraphics[width = \textwidth]{./graphics/it02-uti.pdf}
\end{figure}

```{r it02, eval=FALSE, include=FALSE}
it02 <- filter(splines, speaker == "it02", label %in% c("max_TT", "max_TD"), fan_line > 5, fan_line < 40) # filter out splines containing spurious data

it02_polar_gam <- polar_gam(
  Y ~
    c2_phonation +
    vowel +
    c2_place +
#    vow_phon +
#    c2_place_phon +
    s(X, bs = "cr") +
    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
#    s(X, by = vow_phon, bs = "cr") +
#    s(X, by = c2_place_phon, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = it02,
  method = "ML"
)

plot_polar_smooths(
  it02_polar_gam,
  X,
  c2_phonation,
  facet_terms = c2_place + vowel
) +
  labs(title = "IT02")

it02_polar_gam_null <- polar_gam(
  Y ~
#    c2_phonation +
    vowel +
    c2_place +
#    vow_phon +
#    c2_place_phon +
    s(X, bs = "cr") +
#    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
#    s(X, by = vow_phon, bs = "cr") +
#    s(X, by = c2_place_phon, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = it02,
  method = "ML"
)

compareML(it02_polar_gam_null, it02_polar_gam)

acf_resid(it02_polar_gam, split_pred = list(it02$rec_date))
```

```{r it03, eval=FALSE, include=FALSE}
it03 <- filter(splines, speaker == "it03", label %in% c("max_TT", "max_TD"), fan_line < 40, vowel == "a")

it03_polar_gam <- polar_gam(
  Y ~
    c2_phonation +
#    vowel +
    c2_place +
#    vow_phon +
#    c2_place_phon +
    s(X, bs = "cr") +
    s(X, by = c2_phonation, bs = "cr") +
#    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
#    s(X, by = vow_phon, bs = "cr") +
#    s(X, by = c2_place_phon, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = it03,
  method = "ML"
)

rho <- start_value_rho(it03_polar_gam)

it03_polar_gam_ar <- polar_gam(
  Y ~
    c2_phonation +
#    vowel +
    c2_place +
    s(X, bs = "cr") +
    s(X, by = c2_phonation, bs = "cr") +
#    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = it03,
  method = "ML",
  rho = rho,
  AR_start = it03$start.event
)


it03_polar_gam_ar_null <- polar_gam(
  Y ~
#    c2_phonation +
#    vowel +
    c2_place +
    s(X, bs = "cr") +
#    s(X, by = c2_phonation, bs = "cr") +
#    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = it03,
  method = "ML",
  rho = rho,
  AR_start = it03$start.event
)

compareML(it03_polar_gam_ar_null, it03_polar_gam_ar)

acf_resid(it03_polar_gam_ar, split_pred = "AR.start")

plot_polar_smooths(
  it03_polar_gam_ar,
  X,
  c2_phonation,
  facet_terms = c2_place
)
```

```{r it04, eval=FALSE, include=FALSE}
it04 <- filter(splines, speaker == "it04", label %in% c("max_TT", "max_TD"))

it04_polar_gam <- polar_gam(
  Y ~
    c2_phonation +
    vowel +
    c2_place +
#    vow_phon +
#    c2_place_phon +
    s(X, bs = "cr") +
    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
#    s(X, by = vow_phon, bs = "cr") +
#    s(X, by = c2_place_phon, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = it04,
  method = "ML"
)

it04_polar_gam_null <- polar_gam(
  Y ~
#    c2_phonation +
    vowel +
    c2_place +
#    vow_phon +
#    c2_place_phon +
    s(X, bs = "cr") +
#    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
#    s(X, by = vow_phon, bs = "cr") +
#    s(X, by = c2_place_phon, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = it04,
  method = "ML"
)

compareML(it04_polar_gam_null, it04_polar_gam)

acf_resid(it04_polar_gam, split_pred = list(it04$rec_date))
```

# Results: Tongue contours

\begin{figure}
  \includegraphics[width = \textwidth]{./graphics/pl04-uti.pdf}
\end{figure}

```{r pl02, eval=FALSE, include=FALSE}
pl02 <- filter(splines, speaker == "pl02", label %in% c("max_TT", "max_TD"))

pl02_polar_gam <- polar_gam(
  Y ~
    c2_phonation +
    vowel +
    c2_place +
#    vow_phon +
#    c2_place_phon +
    s(X, bs = "cr") +
    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
#    s(X, by = vow_phon, bs = "cr") +
#    s(X, by = c2_place_phon, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = pl02,
  method = "ML"
)

pl02_polar_gam_null <- polar_gam(
  Y ~
#    c2_phonation +
    vowel +
    c2_place +
#    vow_phon +
#    c2_place_phon +
    s(X, bs = "cr") +
#    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
#    s(X, by = vow_phon, bs = "cr") +
#    s(X, by = c2_place_phon, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = pl02,
  method = "ML"
)

compareML(pl02_polar_gam_null, pl02_polar_gam)

plot_polar_smooths(
  pl02_polar_gam,
  X,
  c2_phonation,
  facet_terms = c2_place + vowel
)
```

```{r pl03, eval=FALSE, include=FALSE}
pl03 <- filter(splines, speaker == "pl03", label %in% c("max_TT", "max_TD"))

pl03_polar_gam <- polar_gam(
  Y ~
    c2_phonation +
    vowel +
    c2_place +
#    vow_phon +
#    c2_place_phon +
    s(X, bs = "cr") +
    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
#    s(X, by = vow_phon, bs = "cr") +
#    s(X, by = c2_place_phon, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = pl03,
  method = "ML"
)

pl03_polar_gam_null <- polar_gam(
  Y ~
#    c2_phonation +
    vowel +
    c2_place +
#    vow_phon +
#    c2_place_phon +
    s(X, bs = "cr") +
#    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
#    s(X, by = vow_phon, bs = "cr") +
#    s(X, by = c2_place_phon, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = pl03,
  method = "ML"
)
compareML(pl03_polar_gam_null, pl03_polar_gam)

plot_polar_smooths(
  pl03_polar_gam,
  X,
  c2_phonation,
  facet_terms = c2_place + vowel
)

acf_resid(it02_polar_gam, split_pred = list(it02$rec_date))
```

```{r pl04, eval=FALSE, include=FALSE}
pl04 <- filter(splines, speaker == "pl04", label %in% c("max_TT", "max_TD"))

pl04_polar_gam <- polar_gam(
  Y ~
    c2_phonation +
    vowel +
    c2_place +
#    vow_phon +
#    c2_place_phon +
    s(X, bs = "cr") +
    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
#    s(X, by = vow_phon, bs = "cr") +
#    s(X, by = c2_place_phon, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = pl04,
  method = "ML"
)

pl04_polar_gam_null <- polar_gam(
  Y ~
#    c2_phonation +
    vowel +
    c2_place +
#    vow_phon +
#    c2_place_phon +
    s(X, bs = "cr") +
#    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
#    s(X, by = vow_phon, bs = "cr") +
#    s(X, by = c2_place_phon, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = pl04,
  method = "ML"
)
compareML(pl04_polar_gam_null, pl04_polar_gam)

plot_polar_smooths(
  pl04_polar_gam,
  X,
  c2_phonation,
  facet_terms = c2_place + vowel
) +
  labs(title = "PL04")
```

# Results: Tongue contours

\begin{figure}
  \includegraphics[width = 0.8\textwidth]{./graphics/pl05-uti.pdf}
\end{figure}

```{r pl05, eval=FALSE, include=FALSE}
pl05 <- filter(splines, speaker == "pl05", label %in% c("max_TT", "max_TD"))

pl05_polar_gam <- polar_gam(
  Y ~
    c2_phonation +
    vowel +
    c2_place +
#    vow_phon +
#    c2_place_phon +
    s(X, bs = "cr") +
    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
#    s(X, by = vow_phon, bs = "cr") +
#    s(X, by = c2_place_phon, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = pl05,
  method = "ML"
)

pl05_polar_gam_null <- polar_gam(
  Y ~
#    c2_phonation +
    vowel +
    c2_place +
#    vow_phon +
#    c2_place_phon +
    s(X, bs = "cr") +
#    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
#    s(X, by = vow_phon, bs = "cr") +
#    s(X, by = c2_place_phon, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = pl05,
  method = "ML"
)

compareML(pl05_polar_gam_null, pl05_polar_gam)

plot_polar_smooths(
  pl05_polar_gam,
  X,
  c2_phonation,
  facet_terms = c2_place + vowel
) +
  labs(title = "PL05")

acf_resid(pl05_polar_gam, split_pred = list(pl05$rec_date))
```

# Discussion: Results summary

* **Effect of voicing on vowel duration**
    * Italian: +22 msec
    * Polish: +12 msec
* **Tongue contours**
    * 4 of 8 speakers (IT01, IT02, IT03, PL05) show TRA within closure
* **2 speakers** (IT02, PL05) with stronger VE and greater TRA

# Discussion

* **TRA hypothesis**: Longer vowel duration allows for greater tongue root advancement.
    * Cf. with @halle1967: laryngeal adjustments

* If TRA hypothesis is correct:
    - TRA during the vowel
    - Greater TRA in IT02 and PL05

# Discussion: TRA during the vowel

```{r tra-series-gam, include=FALSE}
tra_series_gam <- bam(
  tra_z ~
    vowel +
    c2_place +
    s(proportion, k = 5) +
    s(proportion, by = vowel, k = 5) +
    s(proportion, by = c2_place, k = 5) +
    s(proportion, speaker, bs = "fs", m = 1, k = 5) +
    s(proportion, speaker_contexts, bs = "fs", m = 1, k = 5),
  data = filter(tra_series, c2_phonation == "voiced")
)
```

```{r tra-series-gam-smooths}
plot_smooths(
  tra_series_gam,
  proportion,
  c2_place,
  exclude_random = FALSE,
  conditions = quos(c2_place == "coronal", vowel == "a")
) + facet_wrap(~ speaker, ncol = 4) +
  labs(
    x = "Proportion of vowel duration (ratio)",
    y = "Tongue root advancement (z-scores)"
  ) +
  theme(legend.position = "none", panel.spacing = unit(1, "lines"))
```

# Discussion: Electromiography [EMG, @raphael1975]

<!--
    * electromiography (EMG)
    * sustained muscular activity after EMG peak in vowels followed by voiced stops
    * onset of consonant activity relative to vowel activity offset
-->

\begin{figure}
  \includegraphics[width = 0.8\textwidth]{./graphics/raphael.pdf}
\end{figure}

# Discussion

* @raphael1975: **sustained muscular activity in vowels followed by voiced consonants**
    * extra time allows more tongue root advancement?

* Is this a gradual (linear) relationship?
    * We might see a **positive correlation between vowel duration and degree of TRA** (but caveat!)

# Discussion: Vowel Duration ~ TRA

\begin{figure}
  \includegraphics[width=0.8\textwidth]{./graphics/tra-vow.png}
\end{figure}

# Conclusion

* Durational and ultrasound data from 8 speakers
    * Stronger VE ~ Greater TRA
    * TRA during the vowel
    * Vowel duration ~ TRA

* Future work
    * More speakers
    * Can the TRA gesture account for durational difference?

# Conclusion

\begin{center}
{\LARGE \textbf{THANK YOU!}}
\end{center}

**Aknowledgements**: Research funded by the School of Arts, Languages and Cultures at the University of Manchester. Thanks to Patrycja Strycharczuk, Ricardo Bermúdez-Otero, and the members of the Phonetics Lab at UoM for their invaluable feedback. All errors remain my own.

# References
